Spatiotemporal models of DO

### 2020-11-14 analysis/tmb-sensor-explore/3-fit-do.Rmd
for DO model: plot covariate effects, residuals, and spatial random fields

```{r setup}
library(tidyverse)
library(gfplot)
library(sdmTMB)
library(gfranges)
```

Model diagnostics function
```{r}
get_diag <- function(m, response = "log_do", variable = "depth_scaled") {
  predictions <- predict(m)
  predictions$residuals <- residuals(m)

  plot_map <- function(dat, column = "est") {
    ggplot(dat, aes_string("X", "Y", colour = column)) +
      geom_point() +
      coord_fixed()
  }

@@

  g <- ggplot(predictions, aes_string("est", response)) +
    geom_point(alpha = 0.2) +
    facet_wrap(~year) +
    coord_fixed() +
    geom_abline()
  print(g)

  g <- ggplot(predictions, aes(est, residuals)) +
    geom_point(alpha = 0.2) +
    geom_smooth()
  print(g)

  g <- ggplot(predictions, aes_string(variable, "residuals")) +
    geom_point(alpha = 0.2) +
    geom_smooth() +
    facet_wrap(~year)
  print(g)

  g <- ggplot(predictions, aes(X, Y, colour = residuals)) +
    geom_point(alpha = 0.2) +
    coord_fixed() +
    scale_color_gradient2() +
    facet_wrap(~year)
  print(g)

@@

# Define and run model
Current best model of DO
```{r eval=FALSE}
do_model_all_DOY <- sdmTMB::sdmTMB(
  log_do ~ 0 + as.factor(year) + temp_scaled + temp_scaled2 + DOY_scaled,
  data = d_trawl1,
  time_varying = ~ 0 + depth_scaled + depth_scaled2,
  time = "year",
  spde = spde,
  family = gaussian(link = "identity"),
  ar1_fields = TRUE,
  include_spatial = TRUE,
  nlminb_loops = 2,
  newton_steps = 1,
  silent = FALSE
)

saveRDS(do_model_all_DOY, file = (here::here(
  paste0(
    "analysis/tmb-sensor-explore/models/model-do-without-wcvi2016b-",
    "800kn", # Sys.Date(),
    ".rds"
  )
)))
```

# Check diagnostics
```{r eval=T}
# do_model_all_DOY <- readRDS((here::here("analysis/tmb-sensor-explore/models/model-do-without-wcvi2016-depth3.rds"))) # 500 knots, poly-3 of depth
# do_model_all_DOY <- readRDS((here::here("analysis/tmb-sensor-explore/models/model-do-without-wcvi2016-depth2.rds"))) # 500 knots, poly-2 of depth
# do_model_all_DOY <- readRDS((here::here("analysis/tmb-sensor-explore/models/model-do-without-wcvi2016-800kn.rds"))) # 800 knots
do_model_all_DOY <- readRDS((here::here("analysis/tmb-sensor-explore/models/model-do-without-wcvi2016b-800kn.rds"))) # add in some WCVI 2016 data
get_diag(do_model_all_DOY, response = "log_do", variable = "depth_scaled")
```

@@

Plot predictions from saved DO predictions file
```{r}
p_nd <- readRDS((here::here("analysis/VOCC/data/predicted-DO-2020-06-20-more2016.rds")))

plot_map_raster <- function(dat, column = "est") {
  ggplot(dat, aes_string("X", "Y", fill = column)) +
    geom_raster() +
    coord_fixed() +
    gfplot::theme_pbs() + theme(axis.title.x = element_blank(), axis.title.y = element_blank())
}

plot_map_raster(p_nd, "do_est") +
  scale_fill_viridis_c(trans = "log") +
  facet_wrap(~year) +
  ggtitle("Prediction (fixed effects + all random effects)")
```

@@

```{r}
omega_s_all <- plot_map_raster(p_nd, "omega_s") +
  ggtitle("Spatial random effects only") +
  scale_fill_gradient2(
    high = "Steel Blue 4",
    low = "Red 3", limits = c(-0.65, 1)
  ) +
  labs(fill = " ") +
  gfplot::theme_pbs() + theme(axis.title.x = element_blank(), axis.title.y = element_blank(), legend.position = c(.82, .75))

omega_s_all
```

@@

```{r}
p_nd1 <- p_nd %>%
  filter(ssid != 16) %>%
  filter(ssid != 4)
st_1n3 <- plot_map_raster(p_nd1, "epsilon_st") +
  ggtitle("Spatiotemporal random effects") +
  facet_wrap(~year) +
  scale_fill_gradient2(
    high = "Steel Blue 4",
    low = "Red 3", limits = c(-0.5, 0.5)
  ) +
  labs(fill = " ") +
  gfplot::theme_pbs() + theme(axis.title.x = element_blank(), axis.title.y = element_blank(), legend.position = c(.85, .25))
st_1n3
```

```{r}
p_nd4 <- p_nd %>% filter(ssid == 4) # %>% filter(year!=2016)
st_4 <- plot_map_raster(p_nd4, "epsilon_st") +
  # ggtitle(" ") +
  facet_wrap(~year) +
  scale_fill_gradient2(
    high = "Steel Blue 4",
    low = "Red 3", limits = c(-0.5, 0.5)
  ) +
  labs(fill = "epsilon") +
  gfplot::theme_pbs() + theme(axis.title.x = element_blank(), axis.title.y = element_blank(), legend.position = "none")
st_4
```

```{r}
p_nd16 <- p_nd %>% filter(ssid == 16) # %>% filter(ssid!=4)
st_16 <- plot_map_raster(p_nd16, "epsilon_st") +
  # ggtitle(" ") +
  facet_wrap(~year) +
  scale_fill_gradient2(
    high = "Steel Blue 4",
    low = "Red 3", limits = c(-0.5, 0.5)
  ) +
  xlim(65, 345) +
  ylim(5800, 6100) +
  gfplot::theme_pbs() + theme(axis.title.x = element_blank(), axis.title.y = element_blank(), legend.position = "none")
st_16
```

>>>>

```{r setup}
library(tidyverse)
library(gfplot)
library(sdmTMB)
library(gfranges)

plot_map_raster <- function(dat, column = "est") {
    ggplot(dat, aes_string("X", "Y", fill = column)) +
    geom_raster() +
    coord_fixed()+
gfplot::theme_pbs() + theme(
  axis.title = element_blank(), 
  axis.ticks = element_blank(),
  axis.text = element_blank()
  )
}
```

Model diagnostics function
```{r}
get_diag <- function(m, response = "log_do", variable = "depth_scaled", colour_var =  "raw_depth") {
  predictions <- predict(m)
  predictions$residuals <- residuals(m)

  plot_map <- function(dat, column = "est") {
    ggplot(dat, aes_string("X", "Y", colour = column)) +
      geom_point() +
      coord_fixed()+
      gfplot::theme_pbs() 
  }

@@

  g <- ggplot(predictions, aes_string("est", response)) +
    geom_point(alpha = 0.2) +
    facet_wrap(~year) +
    coord_fixed() +
    geom_abline()+
    gfplot::theme_pbs() 
  print(g)

  g <- ggplot(predictions, aes(est, residuals)) +
    geom_point(alpha = 0.2) +
    geom_smooth()+
    gfplot::theme_pbs() 
  print(g)

g <- ggplot(predictions, aes_string(variable, "residuals", colour = colour_var)) +
    # geom_point(alpha = 0.2) +
    geom_point(alpha = 0.4, size = 0.7) +    
    geom_smooth(colour="grey", size = 0.7) +
    scale_colour_viridis_c(option = "B", direction = -1, begin = 0, end = 0.7, trans= sqrt) +
    facet_wrap(~year, scales = "free_x")+
    gfplot::theme_pbs() + theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
  # g <- ggplot(predictions, aes_string(variable, "residuals")) +
  #   geom_point(alpha = 0.2) +
  #   geom_smooth() +
  #   facet_wrap(~year)+
  #   gfplot::theme_pbs() 
  print(g)

  g <- ggplot(predictions, aes(X, Y, colour = residuals)) +
    geom_point(alpha = 0.2) +
    coord_fixed() +
    scale_color_gradient2() +
    facet_wrap(~year)+
    gfplot::theme_pbs() 
  print(g)

@@

# Define and run model
Current best model of DO
```{r eval=FALSE}

# do_model_all_DOY <- readRDS((here::here("analysis/tmb-sensor-explore/models/model-do-without-wcvi2016b-800kn.rds")))
d_trawl1 <- do_model_all_DOY$data
spde <- make_mesh(d_trawl1, xy_cols=c("X", "Y"), n_knots = 800, type = "kmeans")
plot(spde)
do_model_all_DOY <- sdmTMB::sdmTMB(
  log_do ~ 0 + as.factor(year) + temp_scaled + temp_scaled2 + DOY_scaled,
  data = d_trawl1,
  time_varying = ~ 0 + depth_scaled + depth_scaled2,
  time = "year",
  spde = spde,
  family = gaussian(link = "identity"),
  ar1_fields = TRUE,
  include_spatial = TRUE,
  nlminb_loops = 2,
  newton_steps = 1,
  silent = FALSE
)

saveRDS(do_model_all_DOY, file = (here::here(
  paste0(
    "analysis/tmb-sensor-explore/models/model-do-without-wcvi2016b-",
    "800kn", # Sys.Date(),
    "-update.rds"
  )
)))
```

# Check diagnostics
```{r eval=T}
# do_model_all_DOY <- readRDS((here::here("analysis/tmb-sensor-explore/models/model-do-without-wcvi2016-depth3.rds"))) # 500 knots, poly-3 of depth
# do_model_all_DOY <- readRDS((here::here("analysis/tmb-sensor-explore/models/model-do-without-wcvi2016-depth2.rds"))) # 500 knots, poly-2 of depth
# do_model_all_DOY <- readRDS((here::here("analysis/tmb-sensor-explore/models/model-do-without-wcvi2016-800kn.rds"))) # 800 knots
do_model_all_DOY_old <- readRDS((here::here("analysis/tmb-sensor-explore/models/model-do-without-wcvi2016b-800kn.rds"))) # add in some WCVI 2016 data
do_model_all_DOY <- readRDS((here::here("analysis/tmb-sensor-explore/models/model-do-without-wcvi2016b-800kn-update.rds"))) # rerun in new format. No other changes
# do_model_all_DOY <- sdmTMB:::update_model(do_model_all_DOY)
get_diag(do_model_all_DOY, response = "log_do", variable = "depth_scaled", colour_var =  "raw_depth")

get_diag(do_model_all_DOY, response = "log_do", variable = "DOY_scaled", colour_var =  "raw_depth")
get_diag(do_model_all_DOY, response = "log_do", variable = "temp_scaled", colour_var =  "raw_depth")
```

@@

Time-varying depth 
```{r eval = F}
# pd <- expand.grid(
#   temp_scaled = 0,
#   temp_scaled2 = 0,
#   DOY_scaled = 0, 
#   depth_scaled = seq(min(do_model_all_DOY$data$depth_scaled),
#     max(do_model_all_DOY$data$depth_scaled), length.out = 30),
#   year = unique(do_model_all_DOY$data$year)
# )
# pd$depth_scaled2 <- pd$depth_scaled^2
# 
# p <- predict(do_model_all_DOY, newdata = pd, se_fit = TRUE, re_form = NA)
# 
# p$DO <- exp(p$est)
# p$lowCI <- exp(p$est - 1.96 * p$est_se)
# p$highCI <- exp(p$est + 1.96 * p$est_se)
# 
# p$Depth <- exp(p$depth_scaled*do_model_all_DOY$data$depth_sd[1]+do_model_all_DOY$data$depth_mean[1])
# saveRDS(p, here::here("analysis/tmb-sensor-explore/data/depth-tv-m-do-DOY-800kn.rds"))

p <- readRDS(here::here("analysis/tmb-sensor-explore/data/depth-tv-m-do-DOY-800kn.rds"))

ggplot(filter(p, year != 2007), aes(Depth, exp(est), 
  ymin = lowCI,
  ymax = highCI,
  group = as.factor(year), fill = year, colour = year)) +
  geom_line(size = 0.5, alpha =0.85) +
  # geom_smooth(method= "loess", size = 0.4, alpha =0.85, se =F ) +
  # geom_smooth(span = 0.5, size = 0.4, alpha =0.85, se =F ) +
  scale_fill_viridis_c(option = "C") +
  scale_colour_viridis_c(option = "C") +
  ylab("DO") +
  geom_ribbon(alpha = 0.2, colour = NA) +
  scale_y_continuous(limits = c(0.5,5)) +
  scale_x_reverse(limits = c(650,20)) +
  coord_flip() +
  # scale_x_continuous(limits = c(18,210)) +
  gfplot::theme_pbs() +
  theme(legend.position = c(.8,.4))
ggsave(filename = here::here("analysis/tmb-sensor-explore/figs/depth-tv-m-do-DOY-800-kn.png"), width = 3, height = 2.5)
```

Temperature effect
```{r}
### temp
min(do_model_all_DOY$data$temp_scaled)
# 
# pd <- expand.grid(
#   temp_scaled = seq(min(do_model_all_DOY$data$temp_scaled) + 0.1,
#     max(do_model_all_DOY$data$temp_scaled) - 0.1, length.out = 20),
#   DOY_scaled = 0, 
#   depth_scaled = 0,
#   depth_scaled2 = 0,
#   # year = c(2017L, 2018L)
#   year = unique(do_model_all_DOY$data$year) # seems to need multiple years? why?
# ) %>% filter(year>2016)
# # pd$depth_scaled2 <- pd$depth_scaled^2
# pd$temp_scaled2 <-  pd$temp_scaled^2
# # p <- predict(do_model_all_DOY, newdata = pd, se_fit = TRUE, re_form = NA, xy_cols = c("X", "Y"))
# p <- predict(do_model_all_DOY, newdata = pd, se_fit = TRUE, re_form = NA)
# 
# p$DO <- exp(p$est)
# p$lowCI <- exp(p$est - 1.96 * p$est_se)
# p$highCI <- exp(p$est + 1.96 * p$est_se)
# 
# 
p$Temperature <- do_model_all_DOY$data$temp_sd[1]*p$temp_scaled+do_model_all_DOY$data$temp_mean[1]
saveRDS(p, here::here("analysis/tmb-sensor-explore/data/temp-m-do-DOY-800-kn.rds"))
p <- readRDS(here::here("analysis/tmb-sensor-explore/data/temp-m-do-DOY-800-kn.rds"))

ggplot(filter(p, Temperature < 15 & year <2018), aes(Temperature, DO, 
  ymin = lowCI, 
  ymax = highCI)) +
  geom_line(size = 1.5, alpha =0.85) + 
  # scale_color_viridis_c(option = "C") +
  geom_ribbon(alpha=0.2) +
  # ylim(20, 210) +
  gfplot::theme_pbs() + 
  theme(legend.position = c(.8,.4))

ggsave(filename = here::here("analysis/tmb-sensor-explore/figs/temp-m-do-DOY-800-kn.png"), width = 3, height = 2.5)

```

Seasonal effect
```{r}
# pd <- expand.grid(
#   temp_scaled = 0,
#   temp_scaled2 = 0,
#   DOY_scaled = seq(min(do_model_all_DOY$data$DOY_scaled) + 0.1,
#     max(do_model_all_DOY$data$DOY_scaled) - 0.1, length.out = 5),
#   depth_scaled = 0,
#   depth_scaled2 = 0,
#   year = unique(do_model_all_DOY$data$year) # seems to need multiple years? why?
# ) 
# 
# p <- predict(do_model_all_DOY, newdata = pd, se_fit = TRUE, re_form = NA)
# 
# p$DO <- exp(p$est)
# p$lowCI <- exp(p$est - 1.96 * p$est_se)
# p$highCI <- exp(p$est + 1.96 * p$est_se)
# 
# p$DOY <- p$DOY_scaled*do_model_all_DOY$data$DOY_sd[1]+ do_model_all_DOY$data$DOY_mean[1]
# 
# saveRDS(p, here::here("analysis/tmb-sensor-explore/data/DOY-do-800-kn.rds"))
p <- readRDS(here::here("analysis/tmb-sensor-explore/data/DOY-do-800-kn.rds"))

ggplot(filter(p), aes(DOY, DO, 
  ymin = lowCI, 
  ymax = highCI, 
  group= as.factor(year), fill = year, colour = year)) +
  geom_line(size = 1.5, alpha =0.85) + 
  scale_fill_viridis_c(option = "C") +
  scale_color_viridis_c(option = "C") +
  geom_ribbon(alpha = 0.05, colour = NA) +
  # ylim(20, 210) +
  coord_cartesian(ylim = c(1,5)) +
  gfplot::theme_pbs()# + 
  # theme(legend.position = c(.8,.7))

ggsave(filename = here::here("analysis/tmb-sensor-explore/figs/do-DOY-800-kn.png"), width = 3, height = 2.5)
```

Plot predictions from saved DO predictions file
```{r}
p_nd <- readRDS((here::here("analysis/VOCC/data/predicted-DO-2020-06-20-more2016.rds")))
p_nd <- readRDS("~/github/dfo/gfranges/analysis/VOCC/data/predicted-DO-2020-06-20-more2016-roms.rds")
plot_map_raster(p_nd, "do_est") +
  scale_fill_viridis_c(trans = "log") +
  facet_wrap(~year) +
  ggtitle("Prediction (fixed effects + all random effects)")
```

@@

```{r}
p_nd <- readRDS("~/github/dfo/gfranges/analysis/VOCC/data/predicted-DO-2020-06-20-more2016-roms.rds")

(omega_s_all <- plot_map_raster(p_nd, "omega_s") +
  ggtitle("Spatial random effects only") +
  scale_fill_gradient2(
    # high = "Steel Blue 4", low = "Red 3",
    high = "gold", low = "darkcyan",
    limits = c(-0.65, 1)
  ) +
  labs(fill = " ") +
  theme(legend.position = c(.2, .3)))
ggsave(filename = here::here("analysis/tmb-sensor-explore/figs/omegas-do-rw-800-kn-yg.png"), width = 3, height = 3)
omega_s_all
```

@@

```{r}
p_nd <- readRDS("~/github/dfo/gfranges/analysis/VOCC/data/predicted-DO-2020-06-20-more2016-roms.rds")

p_nd1 <- p_nd %>%
  filter(ssid != 16) %>%
  filter(ssid != 4)

(st_1n3 <- plot_map_raster(p_nd1, "epsilon_st") +
  ggtitle("Spatiotemporal random effects") +
  facet_grid(~year) +
  scale_fill_gradient2(
    # high = "Steel Blue 4", low = "Red 3", 
    high = "gold", low = "darkcyan",
    limits = c(-0.65, 1)
  ) +
  labs(fill = " ") + theme( legend.position = "none") )

ggsave(filename = here::here("analysis/tmb-sensor-explore/figs/do-epsilon-1n3-yg.png"), width = 6, height = 2)
# st_1n3
```

```{r}
p_nd <- readRDS("~/github/dfo/gfranges/analysis/VOCC/data/predicted-DO-2020-06-20-more2016-roms.rds")

p_nd4 <- p_nd %>% filter(ssid == 4) # %>% filter(year!=2016)
(st_4 <- plot_map_raster(p_nd4, "epsilon_st") +
  # ggtitle(" ") +
  facet_grid(~year) +
  scale_fill_gradient2(
    # high = "Steel Blue 4", low = "Red 3", 
    high = "gold", low = "darkcyan",
    limits = c(-0.65, 1)
  ) +
  labs(fill = "epsilon") + theme( legend.position = "none") )

ggsave(filename = here::here("analysis/tmb-sensor-explore/figs/do-epsilon-4-yg.png"), width = 6, height = 2)
# st_4
```

```{r}
p_nd <- readRDS("~/github/dfo/gfranges/analysis/VOCC/data/predicted-DO-2020-06-20-more2016-roms.rds")

p_nd16 <- p_nd %>% filter(ssid == 16) # %>% filter(ssid!=4)
(st_16 <- plot_map_raster(p_nd16, "epsilon_st") +
  # ggtitle(" ") +
  facet_grid(~year) +
  scale_fill_gradient2(
    # high = "Steel Blue 4", low = "Red 3", 
    high = "gold", low = "darkcyan",
    limits = c(-0.65, 1)
  ) +
  xlim(20,360) + ylim(5850,6050)+ 
  # xlim(65,345) + ylim(5800,6100)+ 
    theme(panel.border = element_blank(), legend.position = "none") )

ggsave(filename = here::here("analysis/tmb-sensor-explore/figs/do-epsilon-16-yg.png"), width = 6, height = 1)
# st_16
```

```{r}
p_nd <- readRDS("~/github/dfo/gfranges/analysis/VOCC/data/predicted-DO-2020-06-20-more2016-roms.rds")

ggplot(p_nd, aes(roms_temp, temp, colour = depth)) +
  geom_point(alpha = 0.1, size = 0.5) + 
  scale_color_viridis_c(option = "B", begin = 0, end = 0.7, direction = -1, trans= sqrt) + 
  xlim(2,15) +
  ylim(2,15) +
  geom_abline(intercept = 0, slope = 1, colour="grey") +
  ylab("Mean May-August temperature from CTD only model ") +
  xlab("Mean April-September temperature from ROMS model ") +
  gfplot::theme_pbs() + 
    theme(legend.position = "none") 
ggsave(filename = here::here("analysis/tmb-sensor-explore/figs/roms.png"), width = 4.5, height =4.5)

```
